Design and Implementation of Probabilistic Programming Language Anglican

### 2016-07-04 paper/paper.tex
style for listings

\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage[lighttt]{lmodern}
\ttfamily
\DeclareFontShape{OT1}{lmtt}{m}{it}
     {<->sub*lmtt/m/sl}{}
	 \usepackage{listings}

@@

An Anglican program, or \textit{query}, is compiled into a Clojure
function. When inference is performed with a provided algorithm, this
produces a sequence of return values, or \textit{predicts}. Anglican
shares a common syntax with Clojure; Clojure functions can be called
from Anglican code and vice versa. A simple program in Anglican can
look like the following code:
\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false]
(defquery models
  "chooses a distribution
   which describes the data"
  (let [;; Model - randomly choose distribution
	    dist (sample (categorical
                       [[uniform-continuous 1]
                        [normal 1] 
                        [gamma 1]]))
        a (sample (gamma 1 1))
        b (sample (gamma 1 1))
        d (dist a b)]
    ;; Data - samples from the distribution
    (observe d 1)
    (observe d 2)
    (observe d 4)
    (observe d 7)
    ;; Output --- predicted distribution
    (predict :d (type d))
    (predict :a a) (predict :b b)))
\end{lstlisting}

@@

To realize potential danger of code size explosion, consider
CPS transformation of code
\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false]
(if (adult? person)
  (if (male? person)
     (choose-beer)
     (choose-wine))
  (choose-juice))
\end{lstlisting}
with continuation 
\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false]
(fn [choice _]
  (case (kind choice)
    :beer (beer-jar choice)
    :wine (wine-glass choice)
    :juice (juice-bottle choice)))
\end{lstlisting}
The code of the continuation, represented by an \texttt{fn}
form, will be repeated three times. In general, CPS code can
grow extremely large if symbolic continuations are inserted
repeatedly.

@@

Providing an explicit syntax for differentiating between
Anglican and Clojure function calls would be cumbersome. Another
option would be to use meta-data to identify Anglican function
calls at runtime, however this would impact performance, and
a good runtime performance is critical for probabilistic
programs. The approach taken by Anglican is to maintain a list
of unqualified names of primitive functions, as well of
namespaces in which all functions are primitive, and recognize
primitive functions by name --- if a function name is not in the
list, the function is a Clojure function. Global dynamically
bound variables \texttt{*primitive-procedures*} and
\texttt{*primitive-namespace*} contain the initial lists of
names and namespaces, correspondingly. Of course, local bindings
can shade global primitive function names. For example,
\texttt{first} is a Clojure function inside the let block in the
following example:
\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false]
(let [first (fn [[x & y]] x)]
  (first '[1 2 3]))
\end{lstlisting}
The Anglican compiler takes care of the shading by rebinding
\texttt{*primitive-procedures*} in every lexical scope
(\texttt{fn-cps}, \texttt{cps-of-let}). Macro
\texttt{shading-primitive-procedures} automates the shading.

@@

The sole purpose of the algorithm identifier is method dispatch
--- conventionally, the identifier is a keyword related to the
algorithm name (\texttt{:lmh} for Lightweight
Metropolis-Hasting, \texttt{:pcascade} for Particle Cascade
etc.). The second parameter is a query as defined by
\texttt{query} or \texttt{defquery} forms. If needed, the query
can be defined anonymously, right in the argument list of a call
to \texttt{infer}:
\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false]
(let [x 1]
  (infer :pgibbs (query (predict x)) nil))
\end{lstlisting}
A query is executed by calling the initial continuation of the
query, which accepts a value and a state. The state is supplied
by the inference algorithm, however the value is provided as a
parameter of \texttt{infer}. A query does not have to have any
parameters, in which case the value can be simply \texttt{nil}.
When a query is defined with a binding for the initial value,
the value becomes available inside the query. The value is
destructured in case of a structured binding, for example:
\begin{lstlisting}[frame=single, language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false]
(defquery my-query [mean sd]
  (predict (sample (normal mean sd))))

@@

The simplest inference algorithm is \textit{importance
sampling}:
\begin{lstlisting}[frame=single, language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false]
(derive ::algorithm
        :anglican.inference/algorithm)
(defmethod infer :importance [_ prog value & {}]
  (letfn
    [(sample-seq []
       (lazy-seq
         (cons
           (:state (exec ::algorithm prog value
                         initial-state))
           (sample-seq))))]
    (sample-seq)))
\end{lstlisting}
\texttt{anglican.importance/{\linebreak[0]}infer} just calls
\texttt{anglican.inference/{\linebreak[0]}exec} and relies on default
implementations of checkpoint handlers. A different inference
algorithm would provide its own implementations of
\texttt{checkpoint} for \texttt{sample}, \texttt{observe}, or
both, as well as invoke \texttt{exec} from an elaborated
conditional control flow. LMH (\texttt{anglican.lmh}) and SMC
(\texttt{anglican.smc}) are examples of inference algorithms
where either \texttt{observe} (SMC) or \texttt{sample} (LMH)
handler is overridden. In addition, SMC runs multiple particles
(program instances) simultaneously, while LMH runs re-runs
programs from an intermediate continuation rather than from the
beginning.

>>>>

\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{listings}
\usepackage[lighttt]{lmodern}
\ttfamily
\DeclareFontShape{OT1}{lmtt}{m}{it}
     {<->sub*lmtt/m/sl}{}
\lstdefinestyle{default}{frame=single,language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false}

@@

An Anglican program, or \textit{query}, is compiled into a Clojure
function. When inference is performed with a provided algorithm, this
produces a sequence of return values, or \textit{predicts}. Anglican
shares a common syntax with Clojure; Clojure functions can be called
from Anglican code and vice versa. A simple program in Anglican can
look like the following code:
\begin{lstlisting}[style=default]
(defquery models
  "chooses a distribution
   which describes the data"
  (let [;; Model - randomly choose distribution
	    dist (sample (categorical
                       [[uniform-continuous 1]
                        [normal 1] 
                        [gamma 1]]))
        a (sample (gamma 1 1))
        b (sample (gamma 1 1))
        d (dist a b)]
    ;; Data - samples from the distribution
    (observe d 1)
    (observe d 2)
    (observe d 4)
    (observe d 7)
    ;; Output --- predicted distribution
    (predict :d (type d))
    (predict :a a) (predict :b b)))
\end{lstlisting}

@@

To realize potential danger of code size explosion, consider
CPS transformation of code
\begin{lstlisting}[style=default]
(if (adult? person)
  (if (male? person)
     (choose-beer)
     (choose-wine))
  (choose-juice))
\end{lstlisting}
with continuation 
\begin{lstlisting}[style=default]
(fn [choice _]
  (case (kind choice)
    :beer (beer-jar choice)
    :wine (wine-glass choice)
    :juice (juice-bottle choice)))
\end{lstlisting}
The code of the continuation, represented by an \texttt{fn}
form, will be repeated three times. In general, CPS code can
grow extremely large if symbolic continuations are inserted
repeatedly.

@@

Providing an explicit syntax for differentiating between
Anglican and Clojure function calls would be cumbersome. Another
option would be to use meta-data to identify Anglican function
calls at runtime, however this would impact performance, and
a good runtime performance is critical for probabilistic
programs. The approach taken by Anglican is to maintain a list
of unqualified names of primitive functions, as well of
namespaces in which all functions are primitive, and recognize
primitive functions by name --- if a function name is not in the
list, the function is a Clojure function. Global dynamically
bound variables \texttt{*primitive-procedures*} and
\texttt{*primitive-namespace*} contain the initial lists of
names and namespaces, correspondingly. Of course, local bindings
can shade global primitive function names. For example,
\texttt{first} is a Clojure function inside the let block in the
following example:
\begin{lstlisting}[style=default]
(let [first (fn [[x & y]] x)]
  (first '[1 2 3]))
\end{lstlisting}
The Anglican compiler takes care of the shading by rebinding
\texttt{*primitive-procedures*} in every lexical scope
(\texttt{fn-cps}, \texttt{cps-of-let}). Macro
\texttt{shading-primitive-procedures} automates the shading.

@@

The sole purpose of the algorithm identifier is method dispatch
--- conventionally, the identifier is a keyword related to the
algorithm name (\texttt{:lmh} for Lightweight
Metropolis-Hasting, \texttt{:pcascade} for Particle Cascade
etc.). The second parameter is a query as defined by
\texttt{query} or \texttt{defquery} forms. If needed, the query
can be defined anonymously, right in the argument list of a call
to \texttt{infer}:
\begin{lstlisting}[style=default]
(let [x 1]
  (infer :pgibbs (query (predict x)) nil))
\end{lstlisting}
A query is executed by calling the initial continuation of the
query, which accepts a value and a state. The state is supplied
by the inference algorithm, however the value is provided as a
parameter of \texttt{infer}. A query does not have to have any
parameters, in which case the value can be simply \texttt{nil}.
When a query is defined with a binding for the initial value,
the value becomes available inside the query. The value is
destructured in case of a structured binding, for example:
\begin{lstlisting}[style=default]
(defquery my-query [mean sd]
  (predict (sample (normal mean sd))))

@@

The simplest inference algorithm is \textit{importance
sampling}:
\begin{lstlisting}[style=default]
(derive ::algorithm
        :anglican.inference/algorithm)
(defmethod infer :importance [_ prog value & {}]
  (letfn
    [(sample-seq []
       (lazy-seq
         (cons
           (:state (exec ::algorithm prog value
                         initial-state))
           (sample-seq))))]
    (sample-seq)))
\end{lstlisting}
\texttt{anglican.importance/{\linebreak[0]}infer} just calls
\texttt{anglican.inference/{\linebreak[0]}exec} and relies on default
implementations of checkpoint handlers. A different inference
algorithm would provide its own implementations of
\texttt{checkpoint} for \texttt{sample}, \texttt{observe}, or
both, as well as invoke \texttt{exec} from an elaborated
conditional control flow. LMH (\texttt{anglican.lmh}) and SMC
(\texttt{anglican.smc}) are examples of inference algorithms
where either \texttt{observe} (SMC) or \texttt{sample} (LMH)
handler is overridden. In addition, SMC runs multiple particles
(program instances) simultaneously, while LMH runs re-runs
programs from an intermediate continuation rather than from the
beginning.
